Soss

### 2020-01-01 paper/paper.tex
thautwarm edits

\author{Chad Scherrer}
\orcid{0000-0002-1490-0304}
\affiliation{%
  \institution{RelationalAI}
  % \department{Department of Brain and Cognitive Sciences}
  %\streetaddress{43 Vassar St}
  \city{Seattle}
  \state{WA}
  %\postcode{02139}
  %\country{USA}
}
\email{chad.scherrer@gmail.com}

@@

There are many approaches to building a Probabilistic Programming Language (``PPL''). Soss is distinct from most alternatives in a number of ways:

\begin{itemize}
    \item Model specification in Soss is \emph{declarative}; statements are represented not by the order entered by the user, but by the partial order given by their variable dependencies.
    \item Soss models are typically expressed in terms of \emph{inference combinators} (like ``\jl#For#'') that combine distributions in some way to arrive at a new distribution.
    \item Soss models are \emph{first-class}; models can be passed as arguments to other models, or used in place of distributions within a model.
    \item Soss models are \emph{function-like}; abstractly, a model may be considered a function from its arguments to a joint distribution. In particular, specification of ``observed variables'' is separated from model definition.
    \item Values and distributions in Soss are represented internally as \emph{abstract syntax trees}, keeping the internal representation close to that given by the user for maximum flexibility.
    \item Soss uses \emph{staged compilation} for inference primitives like \jl#rand# and \jl#logpdf#, via novel \emph{generalized generated functions} from \texttt{GeneralizedGenerated.jl}.
    \item Soss supports \emph{model transformations}, functions that take a model and return another model.
    \item Soss supports \emph{symbolic simplifcation}, making it easy to inspect or manipulate a symbolic representation of the log-density, or to use it to generate optimized code.
    \item Soss is \emph{extensible}; users can define new inference primitives or model transformations externally, and use them as if they had been included in Soss.
\end{itemize}

Finally, Soss addresses the well-known challenge of naming things in computer science via Cockney rhyming slang (``sauce pan'' rhymes with ``Stan''). {\bf S}oss is {\bf o}pen-{\bf s}ource {\bf s}oftware.

\section{Modeling in Soss}

Figure~\ref{fig:model} shows a plate diagram and Soss implementation of a simplified two-way ANOVA model. The Soss model $m$ has \emph{arguments} $I$ and $J$. Each line of a model is either a \emph{sample statement} (\jl#v ~ rhs#) or an \emph{assign statement} (\jl#v = rhs#). In either case, \jl#v# must be a valid Julia variable name, and \jl#rhs# must be a valid Julia expression. For a sample statement, \jl#rhs# should also be ``distribution-like'', supporting \jl#logpdf# and/or \jl#rand# methods (depending on the inference algorithm to be used).

\begin{figure}[t]
    \centering

    \begin{subfigure}[b]{0.3\textwidth}
        \[
    \begin{aligned}
        \\ i &\in \{1,\cdots,I\}
        \\ j &\in \{1,\cdots,J\}
        \\ λ &\sim \text{Cauchy}_+(0,1)
        \\ \sigma_I &\sim \text{Normal}_+(0,\lambda)
        \\ \sigma_J &\sim \text{Normal}_+(0,\lambda)
        \\ \alpha_i &\sim \text{Normal}(0,\sigma_I)
        \\ \beta_j &\sim \text{Normal}(0,\sigma_J)
        \\ \hat{y}_{ij} &= \alpha_i + \beta_j
        \\ y_{ij} &\sim \text{Normal}(\hat{y}_{ij}, \sigma)
    \end{aligned}
    \]

    \caption{}
    \end{subfigure}
    \begin{subfigure}[b]{0.3\textwidth}
        \scalebox{0.85}{\begin{tikzpicture}

@@

        \end{tikzpicture}}
        \caption{}
    \end{subfigure}
    \hspace{0.1in}
    \begin{subfigure}[b]{0.3\textwidth}

@@

    \caption{A generative model (a), plate diagram (b), and Soss model (c) for a two-way analysis of variance model. Note the specification that $y$ will later be observed is not part of the model, but is instead given at inference time.}
    \label{fig:model}
\end{figure}

\subsection{Joint Distributions}

\subsection{Inference Primitives}

@@

At this writing, the following inference primitive are available:

\begin{description}  
    \item[\texttt{rand}] draws a random sample from a given joint distribution 
    \item[\texttt{particles}] uses \emph{systematic sampling}
    \item[\texttt{logpdf}]
    \item[\texttt{symlogpdf}] 
    \item[\texttt{weightedSample}] 
    \item[\texttt{xform}] 
\end{description}

\subsection{Model Transformations}

\subsubsection*{Conditional predictive models}

\subsubsection*{Causal interventions (``Do'' operator)}

\subsubsection*{Markov blankets}

@@

\subsection{Symbolic Manipulation}

\subsubsection*{Symbolic Log-density}

\subsubsection*{Code Generation}

@@

Besides the purely static code generation via regular macros happening in parsing time, Soss heavily uses a mechanism of "zero-cost" runtime code generation originated by Julia's generated functions \cite{bezanson2012julia}, a.k.a staged functions in the original paper and earlier versions of Julia Programming Language.

@@

To address this, we introduced the works of easing the restrictions of Julia generated functions,
\todo[inline]{cite GG here?}
which allow us to generate closures from generated functions, and make the programs sufficiently powerful.

@@

  function tolc(expr)
    @match expr begin
      :($f($arg)) => App{tolc(f), tolc(arg)}
      :($x => $y) => Lam{x, tolc(y)}
      a::Symbol => Var{a}
    end
  end

  @gg function lc(::Type{Term}) where Term <: LCTerm
      cg(Term)
  end

  macro lc(term)
    lc(tolc(term))
  end

  x = Core.Box()
  y = Core.Box()
  z = Core.Box()
  f(x) = (x, z)
  @assert (@lc x => x)(x) === (@lc y => y)(x)
  @assert (@lc f => x => f(x))(f)(y) === (y, z)
  \end{minted}

>>>>

\author{Chad Scherrer}
\orcid{0000-0002-1490-0304}
\affiliation{%
  \institution{RelationalAI}
  % \department{Department of Brain and Cognitive Sciences}
  %\streetaddress{43 Vassar St}
  %\city{Cambridge}
  %\state{MA}
  %\postcode{02139}
  %\country{USA}
}
\email{chad.scherrer@gmail.com}

@@

There are many approaches to building a Probabilistic Programming Language (``PPL''). Soss is distinct from most in a number of ways...

\section{Model syntax}

\begin{figure}[H]
    \centering

    \begin{subfigure}[b]{0.4\textwidth}
        \begin{tikzpicture}

@@

        \end{tikzpicture}
        \caption{}
    \end{subfigure}
    \hspace{0.1in}
    \begin{subfigure}[b]{0.4\textwidth}

@@

    \caption{A plate diagram (a) and Soss encoding (b) for a two-way analysis of variance model. Note the specification that $y$ will later be observed is not part of the model, but is instead given at inference time.}
    \label{fig:model}
\end{figure}

Figure~\ref{fig:model} shows a plate diagram and Soss implementation of a simplified two-way ANOVA model. The model $m$ has \emph{arguments} $I$ and $J$. Each line of a model is either a \emph{sample statement} (\jl#v ~ rhs#) or an \emph{assign statement} (\jl#v = rhs#). In either case, \jl#v# must be a valid Julia variable name, and \jl#rhs# must be a valid Julia expression. For a sample statement, \jl#rhs# should also be ``distribution-like'', supporting \jl#logpdf# and/or \jl#rand# methods (depending on the inference algorithm to be used).

\section{Inference Primitives}

@@

\begin{figure}

    \begin{minted}{julia}
        julia> truth = rand(m(I=2,J=3)); pairs(truth)
        pairs(::NamedTuple) with 10 entries:
          :I    => 2
          :J    => 3
          :yhat => [0.0996722 0.720247 -0.706772; 0.100505 0.72108 -0.705939]
          :σ    => 0.297288
          :λ    => 0.523386
          :σJ   => 0.312794
          :σI   => 0.0054669
          :β    => [0.0973136, 0.717889, -0.70913]
          :α    => [0.00235854, 0.00319107]
          :y    => [0.386041 0.564963 -0.72177; 0.236898 0.842491 -0.912154]
    \end{minted}

\end{figure}

\section{Model Transformations}

\subsection{Conditional Predictive models}

\subsection{Causal Interventions (``Do'' operator)}

\subsection{Markov Blankets}

@@

\section{Symbolic Manipulation}

\subsection{Symbolic Log-density}

\subsection{Code Generation}

@@

Besides the purely static code generation via regular macros happening in parsing time, Soss heavily uses a mechanism of "zero-cost" runtime code generation originated by Julia's generated functions, a.k.a staged functions in the original paper and earlier versions of Julia Programming Language \cite{bezanson2012julia}.

@@

To address this, we ease the restrictions of Julia generated functions by leveraging GeneralizedGenerated \cite{taine_zhao_2020_3596233} to,
which allows us to generate closures from generated functions.

@@

  @gg lc(::Type{Term}) where Term <: LCTerm = cg(Term)
\end{minted} 
  % tolc(expr::Expr) =  @match expr begin
  %                       :($f($arg)) => App{tolc(f), tolc(arg)}
  %                       :($x => $y) => Lam{x, tolc(y)}
  %                     end
  % tolc(sym::Symbol) => Var{sym}
  % macro lc(term)
  %   lc(tolc(term))
  % end

  % x = Core.Box()
  % y = Core.Box()
  % z = Core.Box()
  % f(x) = (x, z)
  % @assert (@lc x => x)(x) === (@lc y => y)(x)
  % @assert (@lc f => x => f(x))(f)(y) === (y, z)
